% CREATED BY DAVID FRISK, 2016
\thispagestyle{plain}
\section*{Acknowledgements}
I would like to start by thanking my supervisor Aarne Ranta, not only for giving me an opportunity to work on a topic that puts together so many of my interests, but also for the constant dialogue we had throughout the project. \smallskip

For their time and valuable feedback, I am also grateful to my examiner Carl-Johan Seger, my opponent Tarik Ala Hadi and Marie-Philippe Gill. With her I had an informal but prolonged exchange of views on our respective thesis projects. \smallskip

I also want to express my gratitude to the friends who helped me with the translations necessary for one of this thesis' experiments: Michael, Haidar and both Claudios. \smallskip

Furthermore, I want to thank the Museo Nazionale Etrusco di Villa Giulia for letting me use a reproduction of the Pyrgi Tablets as the cover image for this report. That of the Tablets is an early example of parallel text that, for a series of coincidences, kept coming back to my mind during the six months I dedicated to this thesis. In particular, I am thankful to Alessia Argento for taking the time to look for the image based on my vague description. \smallskip

There is also a long list of people I would like to mention despite not being directly involved in the project. 
First of all, I have to thank Claudio once again, both for encouraging me and for putting up with living together during such a busy time (we would say ``\textit{per avermi supportato e sopportato}'').
I also want to thank my family for supporting my sudden decision to pursue my studies in Sweden.
Finally, the list would easily grow too long if I mentioned all the close and distant friends that made my life enjoyable during this time, but I cannot avoid mentioning Herbert, with whom I share a love for languages and Computer Science, and Quelli della B1, the irreplaceable group of people I went through my Bachelor's degree in Perugia with, that still continues being such an important part of my daily life and thoughts.

\vspace{1.5cm}
\hfill
Arianna Masciolini, Gothenburg, \monthname \space \the\year

\newpage				% Create empty back of side
\thispagestyle{empty}
\mbox{}